---
title: "Day 21 Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
library(memoise)
library(ggplot2)
library(reshape2)
library(knitr)
library(dplyr)
library(stringr)
library(tidyverse)
library(readr)
library(collections)
options(scipen = 999)
```


So, it goes
depressurized robot > radiation robot > frozen robot > me

## Part 1
These are the dictionaries to show how to get from one key to another - starting with the digital keypad
Assuming it will always, always be faster to hit the same button twice than to come back to it later

```{r}
d2d<-dict()
d2d$set("^^","A")
d2d$set("^<","v<A")
d2d$set("^>",c("v>A",">vA"))
d2d$set("^v","vA")
d2d$set("^A",">A")
d2d$set("<^",">^A")
d2d$set("<<","A")
d2d$set("<>",">>A")
d2d$set("<v",">A")
d2d$set("<A",">>^A")
d2d$set(">^",c("<^A","^<A"))
d2d$set("><","<<A")
d2d$set(">>","A")
d2d$set(">v","<A")
d2d$set(">A","^A")
d2d$set("v^","^A")
d2d$set("v<","<A")
d2d$set("v>",">A")
d2d$set("vv","A")
d2d$set("vA",c(">^A","^>A"))
d2d$set("A^","<A")
d2d$set("A<","v<<A")
d2d$set("A>","vA")
d2d$set("Av",c("v<A","<vA"))
d2d$set("AA","A")
```


And this is for the numeric keypad.  If it is in here, it is in my data or one of the sample/test data sets.

```{r}
d2n<-dict()
d2n$set("02","^A")
d2n$set("05","^^A")
d2n$set("08","^^^A")
d2n$set("0A",">A")
d2n$set("13",">>A")
d2n$set("15",c("^>A",">^A"))
d2n$set("17","^^A")
d2n$set("1A",c(">>vA"))
d2n$set("1A",c(">>vA"))
d2n$set("29",c(">^^A","^^>A"))
d2n$set("3A","vA")
d2n$set("37",c("^^<<A","<<^^A"))
d2n$set("45",">A")
d2n$set("4A",">>vvA")
d2n$set("56",">A")
d2n$set("59",c("^>A",">^A"))
d2n$set("5A",c("vv>A",">vvA"))
d2n$set("61",c("v<<A","<<vA"))
d2n$set("64","<<A")
d2n$set("67",c("^<<A","<<^A"))
d2n$set("68",c("^<A","<^A"))
d2n$set("6A","vvA")
d2n$set("71","vvA")
d2n$set("75",c("v>A",">vA"))
d2n$set("79",">>A")
d2n$set("80","vvvA")
d2n$set("89",">A")
d2n$set("8A",c("vvv>A",">vvvA"))
d2n$set("94",c("<<vA","v<<A"))
d2n$set("96","vA")
d2n$set("98","<A")
d2n$set("9A","vvvA")
d2n$set("A0",c("<A"))
d2n$set("A1","^<<A")
d2n$set("A3","^A")
d2n$set("A2",c("^<A","<^A"))
d2n$set("A4","^^<<A")
d2n$set("A6","^^A")
d2n$set("A8",c("^^^<A","<^^^A"))
d2n$set("A9","^^^A")
```


First, find all the possible paths for the depressurized robot - because all of these are the same length, this should just grab the info

```{r}
nopressure<-function(numberpad,code){
  #initialize the code with "A"
  code<-str_c("A",code)
  q<-queue()
  out<-c()
  q$push(c(code,""))
  while(q$size()>0){
    cde<-q$pop()
    #### if no more characters, enter the directional keypad sequence into out
    if(nchar(cde[1])==1){
      out<-c(out,cde[2])
      next}
    ### otherwise, use the dictionary to find sequences
    x<-numberpad$get(str_sub(cde[1],end=2))
    ### for each possible path in the dictionary,
    ### push the code without the first character and add the moves to the end of the sequence of moves 
    for(i in 1:length(x)){
      q$push(c(str_sub(cde[1],start=2),str_c(cde[2],x[i])))}}
out}


```



```{r}
r1<-nopressure(d2n,"029A")
r1
```
Then find the path for the radiation robots

```{r}
radiation<-function(dpad,codes){
  q<-queue()
  #### smallest length
  shortest<-Inf
  smallest<-list()
  for(i in 1:length(codes)){
    x<-codes[i]
    ### pushing the current, the output, and the starting
    q$push(c(str_c("A",x),"",x))}
    while(q$size()>0){
      cde<-q$pop()
      ### check to see if this has worked its way through the sequence
      if(nchar(cde[1])==1){
        ### if this is the shortest, update everything
        if(nchar(cde[2])<shortest){
          shortest<-nchar(cde[2])
          smallest<-list(cde[2:3])
          ### if it is the same length as the shortest - keep it
        }else if(nchar(cde[2])==shortest){
          smallest<-c(smallest,list(cde[2:3]))}
        next}
      #### otherwise, start constructing the 
      x<-dpad$get(str_sub(cde[1],end=2))
      ### for each possible path in the dictionary,
      ### push the code without the first character and add the moves to the end of the sequence of moves 
      for(i in 1:length(x)){
        q$push(c(str_sub(cde[1],start=2),str_c(cde[2],x[i]),cde[3]))}}     
    smallest}
```


```{r}
r2<-radiation(d2d,r1)
```
Then find the paths for the frozen robots (The assumption is that the shorter the path earlier, the shorter the path later)

```{r}
frozen<-function(dpad,codes){
  pq<-priority_queue()
  #### smallest length
  shortest<-Inf
  smallest<-list()
  for(i in 1:length(codes)){
    x<-codes[[i]]
    ### pushing the current, the output, and the 
    pq$push(c(str_c("A",x[1]),"",x[1:2]),-nchar(x[1]))}
    while(pq$size()>0){
      cde<-pq$pop()
      ### check to see if this has worked its way through the sequence
      if(nchar(cde[2])>shortest){next}
      if(nchar(cde[1])==1){
        ### if this is the shortest, update everything
        if(nchar(cde[2])<shortest){
          shortest<-nchar(cde[2])
          smallest<-list(cde[-1])
          ### if it is the same length as the shortest - keep it
        }else if(nchar(cde[2])==shortest){
          smallest<-c(smallest,list(cde[-1]))}
        next}
      #### otherwise, start constructing the 
      x<-dpad$get(str_sub(cde[1],end=2))
      ### for each possible path in the dictionary,
      ### push the code without the first character and add the moves to the end of the sequence of moves 
      for(i in 1:length(x)){
        pq$push(c(str_sub(cde[1],start=2),str_c(cde[2],x[i]),cde[-c(1:2)]),-(nchar(cde[1])+nchar(cde[2])+nchar(x[i])))}}     
    shortest}
```


```{r}
r3<-frozen(d2d,r2)
r3
```
For part 1, run through each function and then multiply by the numerical values in the initial code.

```{r}
part1<-sapply(c("029A","980A","179A","456A","379A"),function(x){
  r1<-nopressure(d2n,x)
  r2<-radiation(d2d,r1)
  r3<-frozen(d2d,r2)
  nums<-as.numeric(str_flatten(unlist(str_extract_all(x,"[:digit:]"))))
  cplx<-r3*nums
  cat(x,"-",length(r1),"-",length(r2),"-",r3,"-",cplx,"\n")
  cplx})
sum(part1)
```

### Part 2

Obviously this is too much to do 26 times.  So recursively


```{r}
findshortest<-function(longcode,dct,rep){
  ####if the only code is "A" return 0 
  if(longcode=="A"){return(0)
    ### if there are only two characters, return the number of keys you need to press to get from that character to the next
  }else if(nchar(longcode)==2){
    return(memdbl(longcode,dct,rep))
    ### otherwise, return the sum of the first two characters and the number of keys to press for the remainder of the string. 
  }else{
    return(sum(memdbl(str_sub(longcode,end=2),dct,rep),
               findshortest(str_sub(longcode,start=2)
                            ,dct
                            ,rep)))}}

dblshort<-function(pr,dct,rep){
  ### if this is the final repeat, return the number of characters for that key sequence
  if(rep==1){
    return(min(nchar(dct$get(pr))))
  }else{
    ### otherwise, get all of the options for that key sequence
    ### for each of them, run findshortest() with one fewer repeat
    ### take the minimum.
    return(min(sapply(dct$get(pr),function(x){findshortest(str_c("A",x),dct,rep-1)})))}}

### memoise dbleshort - because otherwise this takes too long
memdbl<-memoise(dblshort)

```

With that in place, use those, plus the nopressure() from above (for the digital keypad) and multiply it by the numeric value.

```{r}
howcomplex<-function(code,ndict,ddict,reps){
  ### first find how the depressurized robot needs to move and 
  pressureless<-str_c("A",nopressure(ndict,code))
  fewestpresses<-min(sapply(pressureless,function(x){findshortest(x,ddict,reps)}))
  nums<-as.numeric(str_flatten(unlist(str_extract_all(code,"[:digit:]"))))
  fewestpresses*nums}
```


```{r}
p2<-sapply(c("029A","980A","179A","456A","379A"),function(x){howcomplex(x,d2n,d2d,25)})
part2<-sum(p2)
part2
```

