---
title: "Day 17 Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
library(ggplot2)
library(reshape2)
library(knitr)
library(dplyr)
library(stringr)
library(tidyverse)
library(readr)
library(collections)
options(scipen = 999)
```


```{r}
programcode<-sapply(str_split(read_lines("../../AoCData/AOC2024/Day17_1.txt"), ","),as.numeric)
startingreg<-sapply(str_split(read_lines("../../AoCData/AOC2024/Day17_2.txt"), ","),as.numeric)
```


```{r}
### regs is the vector in question
regs<-vector("integer",3)
```

```{r,echo=FALSE}
xorbit<-function(a,b){
  bina<-c()
  for(x in 63:0){
    bina<-c(bina,a%/%(2^x))
    a<-a%%(2^x)}

  binb<-c()
  for(x in 63:0){
    binb<-c(binb,b%/%(2^x))
    b<-b%%(2^x)}
  
  c<-as.numeric(xor(bina,binb))

  d<-0
  for(i in 1:length(c)){
    d<-2*d+c[i]}
d}
```

## Part 1

Simulate the computer

```{r}
run_ops<-function(insts,r){
  i<-1
  out<-c()
  while(i<=length(insts)){
    ### get the operands
    litop<-insts[i+1]
    c<-as.character(insts[i+1])
    cop<-switch(c,
                "0"=0,
                "1"=1,
                "2"=2,
                "3"=3,
                "4"=r[1],
                "5"=r[2],
                "6"=r[3],
                "7"={},
                cat("bad operand\n"))
    opchar<-as.character(insts[i])
    switch(opchar,
           ###  A<-A%/%2^cop
           "0"={r[1]<-r[1]%/%(2^cop)},
           ### B<-bitwOr(B,litop)
           ### xorbit is defined later as part of part 2. It's a bitwise xor
           "1"={r[2]<-xorbit(r[2],litop)},
           ### B<-cop%%8
           "2"={r[2]<-cop%%8},
           ### if A!=0 - JUMP to the literal operand (adding one for the offset)
           "3"={
             if(r[1]!=0){
               i<-litop+1
               next}},
           ### B<-bitwOr(B,C)
           "4"={r[2]<-xorbit(r[2],r[3])},
           ### output
           "5"={out<-c(out,cop%%8)},
           ### B<-B%/%2^cop
           "6"={r[2]<-r[1]%/%(2^cop)},
           "7"={r[3]<-r[1]%/%(2^cop)},
           cat(opchar,"bad opcode\n"))
    i<-i+2}
  out}
```

```{r}
p1<-run_ops(programcode,startingreg)
part1<-str_c(p1,collapse=",")
part1
```
## Part 2

Need to break this down - 

My input is 16 numbers long with one output instruction and jumps at the end if A!=0
A=Div(A,8) once per run, so A must be between 8^15 and 8^16-1

And because those numbers are too big, it can't use bitwXor()

```{r}
xorbit<-function(a,b){
  bina<-c()
  for(x in 63:0){
    bina<-c(bina,a%/%(2^x))
    a<-a%%(2^x)}

  binb<-c()
  for(x in 63:0){
    binb<-c(binb,b%/%(2^x))
    b<-b%%(2^x)}
  
  c<-as.numeric(xor(bina,binb))

  d<-0
  for(i in 1:length(c)){
    d<-2*d+c[i]}
d}
```

My instructions are:
B=A%%8\\
B=xor(B,5)\\    
C=div(A,(2^B))\\
B=xor(B,6)\\
A=div(A,8)\\
B=xor(B,C)\\
print B mod 8\\
return to the top\\

So, B&C get reset every run (so they do not matter) only A does. Every run, A drops its last three bits and doesn't change in any other way. And so it can be worked backwards, the first three bits of A determine the last output, the second three bits determine the second to last, & so on.

So:
```{r}
### starting at 0
finala<-c(0)
### for each number in the target
for(i in 1:16){
  tmp<-c()
  ### for every possible a
  for(j in 1:length(finala)){
    maybea<-finala[j]
    ### try adding 0 through 7 to 8*that number (or try with the next three bits) and get the first output
    check<-sapply(0:7,function(x){(run_ops(programcode,c(maybea*8+x,0,0)))[1]})
    ### find which of those is equal to the target output
    ok<-which(check==target[17-i])
    ### if any of them are
    if(length(ok)>0){
      ### adjust because it is indexed by 1
      ok<-ok-1
      ### add that to the potential a*8
      tmp<-c(tmp,maybea*8+ok)}}
    ### update the trials for another run
    finala<-tmp}
### the answer is the lowest possible value
  min(finala)
```

```{r,include=FALSE,eval=FALSE}
### this is a rewrite of the program to ensure I knew what it did.  
rewrite<-function(a,b,c){
  out<-c()
  while(a>0){   #3,0
  b<-a%%8       #2,4
  b<-xorbit(b,5)#1,5
  c<-a%/%(2^b)  #7,5
  b<-xorbit(b,6)#1,6
  a<-a%/%8      #0,3
  b<-xorbit(b,c)#4,3
  out<-c(out,b%%8)  #5,5
  }             #3,0
  cat(c,"\t")
  out}
```


