---
title: "Day 20 Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
library(ggplot2)
library(reshape2)
library(knitr)
library(dplyr)
library(stringr)
library(tidyverse)
library(readr)
library(collections)
options(scipen = 999)
```

```{r}
#input<-read_lines("Day20Sample.txt")
input<-read_lines("../../AoCData/AOC2024/Day20.txt")

racetrack<-dict()
### Put valid spaces into a dictionary
for(i in 1:length(input)){
  x<-unlist(str_split(input[i],""))
  for(j in 1:length(x)){
    if(x[j]!="#"){
      racetrack$set(str_flatten(c(i,j),"~"),0)}
    if(x[j]=="S"){racestart<-c(i,j)}
    if(x[j]=="E"){raceend<-c(i,j)}}}
maxy<-length(input)
maxx<-length(x)
```


## Part 1
You'd think I'd learn by now that not everything is a search... and yet.

When someone cheats through a wall, note the time they emerged from the wall - when (later) the non-cheating person passes that place, record how long it took from the moment of cheating until then.
Tally those.

```{r}
cheatercheater<-function(trk,sy,sx,ey,ex,maxy,maxx){
  beenthere<-dict()
  cheattime<-dict()
  cheatbonus<-c()
  pq<-priority_queue()
  ## tracing y,x,steps
  pq$push(c(sy,sx,0),0)
  racetime<-c()
  while(pq$size()>0){
    curr<-pq$pop()
    ### been there?
    if(beenthere$has(str_flatten(curr[1:2],"~"))){next}
    beenthere$set(str_flatten(curr[1:2],"~"),0)
    ### find out if this is a cheating exit
    if(cheattime$has(str_flatten(curr[1:2],"~"))){
      ### if so, figure out your bonus and add that to the cheatbonus list
      cb<-curr[3]-cheattime$get(str_flatten(curr[1:2],"~"))
      cheatbonus<-c(cheatbonus,cb)}
    ### if at the end - return all of the cheat bonuses
    if(all(curr[1:2]==c(ey,ex))){
      cat(curr[3]," here\n")
      return(cheatbonus)}
    #### Otherwise, check all 4 directions.  In each direction,
    ##### if there isn't a wall, go there.
    ##### if there is a wall,  for each direction out of the wall, put the time you would have cheated to that point
    ##up
    if(curr[1]>1){
      if(trk$has(str_flatten(c(curr[1]-1,curr[2]),"~"))){
        pq$push(c(curr[1]-1,curr[2],curr[3]+1),-sum(1,curr[3],abs(c(ey,ex)-c(curr[1]-1,curr[2]))))
      }else{
          cheattime$set(str_flatten(c(curr[1]-2,curr[2]),"~"),c(cheattime$get(str_flatten(c(curr[1]-2,curr[2]),"~"),default=c()),curr[3]+2))
          cheattime$set(str_flatten(c(curr[1]-1,curr[2]+1),"~"),c(cheattime$get(str_flatten(c(curr[1]-1,curr[2]+1),"~"),default=c()),curr[3]+2))
          cheattime$set(str_flatten(c(curr[1]-1,curr[2]-1),"~"),c(cheattime$get(str_flatten(c(curr[1]-1,curr[2]-1),"~"),default=c()),curr[3]+2))}}
    ##down
    if(curr[1]<maxy){
      if(trk$has(str_flatten(c(curr[1]+1,curr[2]),"~"))){
        pq$push(c(curr[1]+1,curr[2],curr[3]+1),-sum(1,curr[3],abs(c(ey,ex)-c(curr[1]+1,curr[2]))))
      }else{
          cheattime$set(str_flatten(c(curr[1]+2,curr[2]),"~"),c(cheattime$get(str_flatten(c(curr[1]+2,curr[2]),"~"),default=c()),curr[3]+2))
          cheattime$set(str_flatten(c(curr[1]+1,curr[2]+1),"~"),c(cheattime$get(str_flatten(c(curr[1]+1,curr[2]+1),"~"),default=c()),curr[3]+2))
          cheattime$set(str_flatten(c(curr[1]+1,curr[2]-1),"~"),c(cheattime$get(str_flatten(c(curr[1]+1,curr[2]-1),"~"),default=c()),curr[3]+2))}}
    ##left
    if(curr[2]>1){
      if(trk$has(str_flatten(c(curr[1],curr[2]-1),"~"))){
        pq$push(c(curr[1],curr[2]-1,curr[3]+1),-sum(1,curr[3],abs(c(ey,ex)-c(curr[1],curr[2]-1))))
      }else{
          cheattime$set(str_flatten(c(curr[1],curr[2]-2),"~"),c(cheattime$get(str_flatten(c(curr[1],curr[2]-2),"~"),default=c()),curr[3]+2))
          cheattime$set(str_flatten(c(curr[1]-1,curr[2]-1),"~"),c(cheattime$get(str_flatten(c(curr[1]-1,curr[2]-1),"~"),default=c()),curr[3]+2))
          cheattime$set(str_flatten(c(curr[1]+1,curr[2]-1),"~"),c(cheattime$get(str_flatten(c(curr[1]+1,curr[2]-1),"~"),default=c()),curr[3]+2))}}
    ##right
    if(curr[2]<maxx){
      if(trk$has(str_flatten(c(curr[1],curr[2]+1),"~"))){
        pq$push(c(curr[1],curr[2]+1,curr[3]+1),-sum(1,curr[3],abs(c(ey,ex)-c(curr[1],curr[2]+1))))
      }else{
        cheattime$set(str_flatten(c(curr[1],curr[2]+2),"~"),c(cheattime$get(str_flatten(c(curr[1],curr[2]+2),"~"),default=c()),curr[3]+2))
        cheattime$set(str_flatten(c(curr[1]-1,curr[2]+1),"~"),c(cheattime$get(str_flatten(c(curr[1]-1,curr[2]+1),"~"),default=c()),curr[3]+2))
        cheattime$set(str_flatten(c(curr[1]+1,curr[2]+1),"~"),c(cheattime$get(str_flatten(c(curr[1]+1,curr[2]+1),"~"),default=c()),curr[3]+2))}}

}
racetime}
```

```{r}
p1<-cheatercheater(racetrack,racestart[1],racestart[2],raceend[1],raceend[2],maxy,maxx)
table(p1)
#part1<-length(which(p1>=100))
#part1
```


## Part 2
Small change - there are two separate sets of rules, one for while you're cheating, one for the path that hasn't cheated yet.  The same idea, though - time recorded when cheating stops at the place the cheating stops.



```{r}
longcheat<-function(trk,sy,sx,ey,ex,maxy,maxx){
  beenthere<-dict()
  cheattime<-dict()
  cheatpath<-dict()
  cheatbonus<-c()
  pq<-priority_queue()
  ## tracing y,x,cheatstart,cheatstop,cheatsteps,steps
  pq$push(c(sy,sx,-100,-100,0,0),0)
  racetime<-c()
  while(pq$size()>0){
    curr<-pq$pop()
    #### if been cheating too long, stop
    if(curr[5]>19){next}
    ### been there
    if(beenthere$has(str_flatten(curr[1:4],"~"))){next}
    beenthere$set(str_flatten(curr[1:4],"~"),0)
    ### if you have not cheated yet
    if(curr[3]==-100){
      #### find out if this is a space that gets cheated to
      if(cheattime$has(str_flatten(curr[1:2],"~"))){
        ### if so, figure out the bonus and add that to the cheatbonus list
        cb<-curr[6]-cheattime$get(str_flatten(curr[1:2],"~"))
        cheatbonus<-c(cheatbonus,cb)}
      ### if at the end - return all of the cheat bonuses
      if(all(curr[1:2]==c(ey,ex))){
        return(cheatbonus)}
      #### Otherwise, check all 4 directions.  In each direction,
      ##### start cheating in that direction.
      ##### also, with no wall, go there without cheating.
      ##up
      if(curr[1]>1){
        if(trk$has(str_flatten(c(curr[1]-1,curr[2]),"~"))){
          pq$push(c(curr[1]-1,curr[2],-100,-100,0,curr[6]+1),-curr[6])}
          pq$push(c(curr[1]-1,curr[2],curr[1],curr[2],1,curr[6]+1),-curr[6])
          cheatpath$set(str_flatten(c(curr[1]-1,curr[2],curr[1],curr[2]),"~"),0)
          cheattime$set(str_flatten(c(curr[1]-1,curr[2]),"~"),c(cheattime$get(str_flatten(c(curr[1]-1,curr[2]),"~"),default=c()),curr[6]+1))
          }
      ##down
      if(curr[1]<maxy){
        if(trk$has(str_flatten(c(curr[1]+1,curr[2]),"~"))){
          pq$push(c(curr[1]+1,curr[2],-100,-100,0,curr[6]+1),-curr[6])}
          pq$push(c(curr[1]+1,curr[2],curr[1],curr[2],1,curr[6]+1),-curr[6])
          cheatpath$set(str_flatten(c(curr[1]+1,curr[2],curr[1],curr[2]),"~"),0)
          cheattime$set(str_flatten(c(curr[1]+1,curr[2]),"~"),c(cheattime$get(str_flatten(c(curr[1]+1,curr[2]),"~"),default=c()),curr[6]+1))
          }
      ##left
      if(curr[2]>1){
        if(trk$has(str_flatten(c(curr[1],curr[2]-1),"~"))){
          pq$push(c(curr[1],curr[2]-1,-100,-100,0,curr[6]+1),-curr[6])}
          pq$push(c(curr[1],curr[2]-1,curr[1],curr[2],1,curr[6]+1),-curr[6])
          cheatpath$set(str_flatten(c(curr[1],curr[2]-1,curr[1],curr[2]),"~"),0)
          cheattime$set(str_flatten(c(curr[1],curr[2]-1),"~"),c(cheattime$get(str_flatten(c(curr[1],curr[2]-1),"~"),default=c()),curr[6]+1))
      }
      ##right
      if(curr[2]<maxx){
        if(trk$has(str_flatten(c(curr[1],curr[2]+1),"~"))){
          pq$push(c(curr[1],curr[2]+1,-100,-100,0,curr[6]+1),-curr[6])}
          pq$push(c(curr[1],curr[2]+1,curr[1],curr[2],1,curr[6]+1),-curr[6])
          cheatpath$set(str_flatten(c(curr[1],curr[2]+1,curr[1],curr[2]),"~"),0)
          cheattime$set(str_flatten(c(curr[1],curr[2]+1),"~"),c(cheattime$get(str_flatten(c(curr[1],curr[2]+1),"~"),default=c()),curr[6]+1))
          }
      ### if you are cheating, check to see if over 20 steps. If so, too many, stop
      ### In each direction, add to the cheattime ledger *and* keepgoing
    }else{
      if(curr[1]>1){
        if(!cheatpath$has(str_flatten(c(curr[1]-1,curr[2],curr[3],curr[4]),"~"))){
          pq$push(c(curr[1]-1,curr[2],curr[3],curr[4],curr[5]+1,curr[6]+1),-curr[6])
          cheatpath$set(str_flatten(c(curr[1]-1,curr[2],curr[3],curr[4]),"~"),0)
          cheattime$set(str_flatten(c(curr[1]-1,curr[2]),"~"),c(cheattime$get(str_flatten(c(curr[1]-1,curr[2]),"~"),default=c()),curr[6]+1))}}
      ### down
      if(curr[1]<maxy){
        if(!cheatpath$has(str_flatten(c(curr[1]+1,curr[2],curr[3],curr[4]),"~"))){
          pq$push(c(curr[1]+1,curr[2],curr[3],curr[4],curr[5]+1,curr[6]+1),-curr[6])
          cheatpath$set(str_flatten(c(curr[1]+1,curr[2],curr[3],curr[4]),"~"),0)
          cheattime$set(str_flatten(c(curr[1]+1,curr[2]),"~"),c(cheattime$get(str_flatten(c(curr[1]+1,curr[2]),"~"),default=c()),curr[6]+1))}}
      ### left
      if(curr[2]>1){
        if(!cheatpath$has(str_flatten(c(curr[1],curr[2]-1,curr[3],curr[4]),"~"))){
          pq$push(c(curr[1],curr[2]-1,curr[3],curr[4],curr[5]+1,curr[6]+1),-curr[6])
          cheatpath$set(str_flatten(c(curr[1],curr[2]-1,curr[3],curr[4]),"~"),0)
          cheattime$set(str_flatten(c(curr[1],curr[2]-1),"~"),c(cheattime$get(str_flatten(c(curr[1],curr[2]-1),"~"),default=c()),curr[6]+1))}}
      ### right
      if(curr[2]<maxx){
        if(!cheatpath$has(str_flatten(c(curr[1],curr[2]+1,curr[3],curr[4]),"~"))){
          pq$push(c(curr[1],curr[2]+1,curr[3],curr[4],curr[5]+1,curr[6]+1),-curr[6])
          cheatpath$set(str_flatten(c(curr[1],curr[2]+1,curr[3],curr[4]),"~"),0)
          cheattime$set(str_flatten(c(curr[1],curr[2]+1),"~"),c(cheattime$get(str_flatten(c(curr[1],curr[2]+1),"~"),default=c()),curr[6]+1))}}
    }
  }# closing while
}
```

```{r}
p2<-longcheat(racetrack,racestart[1],racestart[2],raceend[1],raceend[2],maxy,maxx)
part2<-length(which(p2>=100))
part2
```
1110467 is too high
982124 is it... YIKES


maybe?

```{r}
longcheat<-function(trk,sy,sx,ey,ex,maxy,maxx){
  beenthere<-dict()
  cheattime<-dict()
  cheatpath<-dict()
  cheatbonus<-c()
  pq<-priority_queue()
  ## tracing y,x,cheatstart,cheatstop,cheatsteps,steps
  pq$push(c(sy,sx,-100,-100,0,0),0)
  racetime<-c()
  while(pq$size()>0){
    curr<-pq$pop()
    #### if been cheating too long, stop
    if(curr[5]>19){next}
    ### been there
    if(beenthere$has(str_flatten(curr[1:4],"~"))){next}
    beenthere$set(str_flatten(curr[1:4],"~"),0)
    ### if you have not cheated yet
    if(curr[3]==-100){
      #### find out if this is a space that gets cheated to
      if(cheattime$has(str_flatten(curr[1:2],"~"))){
        ### if so, figure out the bonus and add that to the cheatbonus list
        cb<-curr[6]-cheattime$get(str_flatten(curr[1:2],"~"))
        cheatbonus<-c(cheatbonus,cb)}
      ### if at the end - return all of the cheat bonuses
      if(all(curr[1:2]==c(ey,ex))){
        return(cheatbonus)}
      #### Otherwise, check all 4 directions.  In each direction,
      ##### start cheating in that direction.
      ##### also, with no wall, go there without cheating.
      ##up
      if(curr[1]>1){
        if(trk$has(str_flatten(c(curr[1]-1,curr[2]),"~"))){
          pq$push(c(curr[1]-1,curr[2],-100,-100,0,curr[6]+1),-sum(curr[6],abs(c(ey,ex)-c(curr[1]-1,curr[2]))))}
          pq$push(c(curr[1]-1,curr[2],curr[1],curr[2],1,curr[6]+1),500-curr[6])
          cheatpath$set(str_flatten(c(curr[1]-1,curr[2],curr[1],curr[2]),"~"),0)
          cheattime$set(str_flatten(c(curr[1]-1,curr[2]),"~"),c(cheattime$get(str_flatten(c(curr[1]-1,curr[2]),"~"),default=c()),curr[6]+1))
          }
      ##down
      if(curr[1]<maxy){
        if(trk$has(str_flatten(c(curr[1]+1,curr[2]),"~"))){
          pq$push(c(curr[1]+1,curr[2],-100,-100,0,curr[6]+1),-sum(curr[6],abs(c(ey,ex)-c(curr[1]+1,curr[2]))))}
          pq$push(c(curr[1]+1,curr[2],curr[1],curr[2],1,curr[6]+1),500-curr[6])
          cheatpath$set(str_flatten(c(curr[1]+1,curr[2],curr[1],curr[2]),"~"),0)
          cheattime$set(str_flatten(c(curr[1]+1,curr[2]),"~"),c(cheattime$get(str_flatten(c(curr[1]+1,curr[2]),"~"),default=c()),curr[6]+1))
          }
      ##left
      if(curr[2]>1){
        if(trk$has(str_flatten(c(curr[1],curr[2]-1),"~"))){
          pq$push(c(curr[1],curr[2]-1,-100,-100,0,curr[6]+1),-sum(curr[6],abs(c(ey,ex)-c(curr[1],curr[2]-1))))}
          pq$push(c(curr[1],curr[2]-1,curr[1],curr[2],1,curr[6]+1),500-curr[6])
          cheatpath$set(str_flatten(c(curr[1],curr[2]-1,curr[1],curr[2]),"~"),0)
          cheattime$set(str_flatten(c(curr[1],curr[2]-1),"~"),c(cheattime$get(str_flatten(c(curr[1],curr[2]-1),"~"),default=c()),curr[6]+1))
      }
      ##right
      if(curr[2]<maxx){
        if(trk$has(str_flatten(c(curr[1],curr[2]+1),"~"))){
          pq$push(c(curr[1],curr[2]+1,-100,-100,0,curr[6]+1),-sum(curr[6],abs(c(ey,ex)-c(curr[1],curr[2]+1))))}
          pq$push(c(curr[1],curr[2]+1,curr[1],curr[2],1,curr[6]+1),500-curr[6])
          cheatpath$set(str_flatten(c(curr[1],curr[2]+1,curr[1],curr[2]),"~"),0)
          cheattime$set(str_flatten(c(curr[1],curr[2]+1),"~"),c(cheattime$get(str_flatten(c(curr[1],curr[2]+1),"~"),default=c()),curr[6]+1))
          }
      ### if you are cheating, check to see if over 20 steps. If so, too many, stop
      ### In each direction, add to the cheattime ledger *and* keepgoing
    }else{
      if(curr[1]>1){
        if(!cheatpath$has(str_flatten(c(curr[1]-1,curr[2],curr[3],curr[4]),"~"))){
          pq$push(c(curr[1]-1,curr[2],curr[3],curr[4],curr[5]+1,curr[6]+1),500-curr[6])
          cheatpath$set(str_flatten(c(curr[1]-1,curr[2],curr[3],curr[4]),"~"),0)
          cheattime$set(str_flatten(c(curr[1]-1,curr[2]),"~"),c(cheattime$get(str_flatten(c(curr[1]-1,curr[2]),"~"),default=c()),curr[6]+1))}}
      ### down
      if(curr[1]<maxy){
        if(!cheatpath$has(str_flatten(c(curr[1]+1,curr[2],curr[3],curr[4]),"~"))){
          pq$push(c(curr[1]+1,curr[2],curr[3],curr[4],curr[5]+1,curr[6]+1),500-curr[6])
          cheatpath$set(str_flatten(c(curr[1]+1,curr[2],curr[3],curr[4]),"~"),0)
          cheattime$set(str_flatten(c(curr[1]+1,curr[2]),"~"),c(cheattime$get(str_flatten(c(curr[1]+1,curr[2]),"~"),default=c()),curr[6]+1))}}
      ### left
      if(curr[2]>1){
        if(!cheatpath$has(str_flatten(c(curr[1],curr[2]-1,curr[3],curr[4]),"~"))){
          pq$push(c(curr[1],curr[2]-1,curr[3],curr[4],curr[5]+1,curr[6]+1),500-curr[6])
          cheatpath$set(str_flatten(c(curr[1],curr[2]-1,curr[3],curr[4]),"~"),0)
          cheattime$set(str_flatten(c(curr[1],curr[2]-1),"~"),c(cheattime$get(str_flatten(c(curr[1],curr[2]-1),"~"),default=c()),curr[6]+1))}}
      ### right
      if(curr[2]<maxx){
        if(!cheatpath$has(str_flatten(c(curr[1],curr[2]+1,curr[3],curr[4]),"~"))){
          pq$push(c(curr[1],curr[2]+1,curr[3],curr[4],curr[5]+1,curr[6]+1),500-curr[6])
          cheatpath$set(str_flatten(c(curr[1],curr[2]+1,curr[3],curr[4]),"~"),0)
          cheattime$set(str_flatten(c(curr[1],curr[2]+1),"~"),c(cheattime$get(str_flatten(c(curr[1],curr[2]+1),"~"),default=c()),curr[6]+1))}}
    }
  }# closing while
}
```

```{r}
p2<-longcheat(racetrack,racestart[1],racestart[2],raceend[1],raceend[2],maxy,maxx)
part2<-length(which(p2>=100))
part2
```