---
title: "Day 24 Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
library(igraph)
library(ggplot2)
library(ggraph)
library(reshape2)
library(knitr)
library(dplyr)
library(stringr)
library(tidyverse)
library(readr)
library(collections)
options(scipen = 999)
```


This reads the starting values into a dictionary and then the gates into a list. Each gate is a vector in the format c(wire1, operator, wire2, wire3).  I also get the highest z value.

```{r}
input<-read_lines("Day24Sample2.txt")

### the values are in a dictionary
wirevalues<-dict()
### the gates/rules are in a list
gates<-list()
highz<- -Inf
blank<-which(input=="")
xvals<-c()
yvals<-c()
for(i in 1:(blank-1)){
  x<-unlist(str_split(input[i],": "))
  wirevalues$set(x[1],as.numeric(x[2]))
  if(str_starts(x[1],"x")){xvals[as.numeric(str_sub(x[1],start=2))+1]<-as.numeric(x[2])}
  if(str_starts(x[1],"y")){yvals[as.numeric(str_sub(x[1],start=2))+1]<-as.numeric(x[2])}}
for(i in (blank+1):length(input)){
    x<-unlist(str_split(input[i]," "))[c(1:3,5)]
    ### look for the highest z
    if(str_starts(x[1],"z")){highz<-max(highz,as.numeric(str_sub(x[1],start=2)))}
    if(str_starts(x[3],"z")){highz<-max(highz,as.numeric(str_sub(x[3],start=2)))}
    if(str_starts(x[4],"z")){highz<-max(highz,as.numeric(str_sub(x[4],start=2)))}
    gates<-c(gates,list(x))}
```


## Part 1

This simply runs through all of the available rules again and again until there's a value for every z.

```{r}
getzs<-function(wv,g,hz){
  ### get the out vector
  zvals<-rep(5,hz+1)
  ### while any of the zs don't have values yet, 
  while(any(zvals==5)){
    i<-1
    ### run through the gates
    while(i<=length(g)){
      wire<-g[[i]]
      if(wv$has(wire[1])&&wv$has(wire[3])){
      otpt<-switch(wire[2],
             "AND"={(wv$get(wire[1])&&wv$get(wire[3]))},
             "OR"={(wv$get(wire[1])||wv$get(wire[3]))},
             "XOR"={(xor(wv$get(wire[1]),wv$get(wire[3])))},
             {cat("unknown\n",unlist(wire),"\n")})
        ### set the gate
        wv$set(wire[4],otpt)
        ### if the output wire is something that begins with z, update the zvector
        otpt<-as.numeric(otpt)
        if(str_starts(wire[4],"z")){zvals[as.numeric(str_sub(wire[4],start=2))+1]<-otpt}
        ### get rid of that instruction
        g<-g[-i]
      }else{i<-i+1}}}
  zdigits<-0
  ### translates z from binary to dec
  for(i in length(zvals):1){
    zdigits<-2*zdigits+zvals[i]}
list(zvals,zdigits)}
```

```{r}
p1<-getzs(wirevalues,gates,highz)
part1<-p1[[2]]
part1
```

## Part 2

Using my actual data, but input the same way.  Also creates x, y, and x+y in decimal form

```{r, echo=FALSE,}
input<-read_lines("../../AoCData/AOC2024/Day24.txt")

### the values are in a dictionary
wirevalues<-dict()
### the gates/rules are in a list
gates<-list()
highz<- -Inf
blank<-which(input=="")
xvals<-c()
yvals<-c()
for(i in 1:(blank-1)){
  x<-unlist(str_split(input[i],": "))
  wirevalues$set(x[1],as.numeric(x[2]))
  if(str_starts(x[1],"x")){xvals[as.numeric(str_sub(x[1],start=2))+1]<-as.numeric(x[2])}
  if(str_starts(x[1],"y")){yvals[as.numeric(str_sub(x[1],start=2))+1]<-as.numeric(x[2])}}
for(i in (blank+1):length(input)){
    x<-unlist(str_split(input[i]," "))[c(1:3,5)]
    ### look for the highest z
    if(str_starts(x[1],"z")){highz<-max(highz,as.numeric(str_sub(x[1],start=2)))}
    if(str_starts(x[3],"z")){highz<-max(highz,as.numeric(str_sub(x[3],start=2)))}
    if(str_starts(x[4],"z")){highz<-max(highz,as.numeric(str_sub(x[4],start=2)))}
    gates<-c(gates,list(x))}

actualzs<-getzs(wirevalues,gates,highz)
zvals<-actualzs[[1]]
zdec<-actualzs[[2]]
### x in decimal form
xdec<-0
  for(i in length(xvals):1){
    xdec<-2*xdec+xvals[i]}
### y in decimal form
ydec<-0
  for(i in length(yvals):1){
    ydec<-2*ydec+yvals[i]}
x_plus_y<-xdec+ydec
ph<-xdec+ydec
x_plus_y_bin<-c()
while(ph>0){
  x_plus_y_bin<-c(x_plus_y_bin,ph%%2)
  ph<-ph%/%2}
```
Put all the gates into a dataframe, because I find them easier to look through than lists
```{r}
gatesdf<-as.data.frame(matrix(nrow=0,ncol=4))
for(i in 1:length(gates)){gatesdf<-rbind(gatesdf,gates[[i]])}
colnames(gatesdf)<-c("a","op","b","c")
```


This shows the values of x, y, x+y and z followed by a graph.

If x+y = z and the graph isn't wonky, the gates and wires are correct.
This function adds x & y and shows the gates in graphical form
```{r,echo=FALSE,warning=FALSE}
cat("x=",xdec,"\n")
cat("y=",ydec,"\n")
cat("x+y=",x_plus_y,"\n")
cat("z=",zdec,"\n")

knitr::kable(rbind(0:45,xs=c(xvals,0),ys=c(yvals,0),added=c(x_plus_y_bin),zs=zvals))




#### graph of the gates
gatemat<-matrix(ncol=2,nrow=0)
for(i in 1:length(gates)){
  g<-gates[[i]]
  gatemat<-rbind(gatemat,c(g[1],g[4]),c(g[3],g[4]))}
gategraph<-graph_from_edgelist(gatemat)
ea<-c()
for(i in 1:length(gates)){ea<-c(ea,gates[[i]][2],gates[[i]][2])}
edge_attr(gategraph)<-list(op=ea)
gategraphplot<-ggraph(gategraph,layout="mds")+
  geom_edge_fan(aes(color=op),
                arrow=arrow(length= unit(1.5, 'mm')))+
  geom_node_label(aes(label=name),size=1.5,repel=TRUE)
gategraphplot
```

The added numbers show an issue around 6.
The graph shows issues around 6, around 11, around 15/16, and around 36


We can also look at the gates in a slightly different form 

```{r}
gatecheck<-function(){

### get lists of all of the different gates
xorvec<-which(gatesdf$op=="XOR")
orvec<-which(gatesdf$op=="OR")
andvec<-which(gatesdf$op=="AND")

### show progression
allxors<-c()

for(i in 0:44){
  ### start with the xNN from 0 to 44
  targetx<-str_c("x",str_pad(i,2,pad=0))
  ### find the gate which combines x & y
  findxy<-c(which(gatesdf$a==targetx),which(gatesdf$b==targetx))
  findxy1<-intersect(findxy,xorvec)
  findxy2<-intersect(findxy,andvec)
  ### show the xor for x&y
  cat(unlist(gatesdf[findxy1,]))
  ### show what it leads to - for everything but x00
  if(i>0){
      ### find the first carry
    xxory<-gatesdf$c[findxy1]
    carry<-c(which(gatesdf$a==xxory),which(gatesdf$b==xxory))
    carry1<-intersect(carry,xorvec)
    cat(" -> ",unlist(gatesdf[carry1,]))
    ### find the second carry
    carry2<-intersect(carry,andvec)
    cat(" // ",unlist(gatesdf[carry2,]))
    ### find the third carry
    targetor<-gatesdf$c[carry2]
    carry3<-c(which(gatesdf$a==targetor),which(gatesdf$b==targetor))
    carry3<-intersect(carry3,orvec)
    cat(" -> ",unlist(gatesdf[carry3,]))
}
  cat("\n")
  if(i>0){cat("\t\t\t\t\t")}
  cat(unlist(gatesdf[findxy2,]),"->\n")
  if(i>0){cat("\n")}
  }
}
gatecheck()

```

From that:

* the outwire from column 1 goes into column 2 and 3.  
* the outwire from column 2 should be zNN
* the outwires from column 3 should be inputs for column 4
* the outwire from column 4 should be inputs for column 2

And also from that:

* z06 & vwr should be switched
* z11 & tqm should be switched
* z16 & kfs should be switched

Something is going on with x36/y36/z36, but it is not as clear as the others.  So starting with the easy to find wires:


```{r}
gateswitcher<-function(a,b,gts){
  for(i in 1:length(gts)){
    if(a==gts[[i]][4]){olda<-i}
    if(b==gts[[i]][4]){oldb<-i}}
  gts[[olda]][4]<-b
  gts[[oldb]][4]<-a
  gts}
gates<-gateswitcher("z06","vwr",gates)
gates<-gateswitcher("z11","tqm",gates)
gates<-gateswitcher("z16","kfs",gates)


gatesdf<-as.data.frame(matrix(nrow=0,ncol=4))
for(i in 1:length(gates)){gatesdf<-rbind(gatesdf,gates[[i]])}
colnames(gatesdf)<-c("a","op","b","c")
```




```{r,echo=FALSE,warning=FALSE}
### recalculate z

actualzs<-getzs(wirevalues,gates,highz)
zvals<-actualzs[[1]]
zdec<-actualzs[[2]]

### re-compare the add & redo the graph

cat("x=",xdec,"\n")
cat("y=",ydec,"\n")
cat("x+y=",x_plus_y,"\n")
cat("z=",zdec,"\n")

knitr::kable(rbind(0:45,xs=c(xvals,0),ys=c(yvals,0),added=c(x_plus_y_bin),zs=zvals))


#### graph of the gates
gatemat<-matrix(ncol=2,nrow=0)
for(i in 1:length(gates)){
  g<-gates[[i]]
  gatemat<-rbind(gatemat,c(g[1],g[4]),c(g[3],g[4]))}
gategraph<-graph_from_edgelist(gatemat)
ea<-c()
for(i in 1:length(gates)){ea<-c(ea,gates[[i]][2],gates[[i]][2])}
edge_attr(gategraph)<-list(op=ea)
gategraphplot<-ggraph(gategraph,layout="mds")+
  geom_edge_fan(aes(color=op),
                arrow=arrow(length= unit(1.5, 'mm')))+
  geom_node_label(aes(label=name),size=1.5,repel=TRUE)
gategraphplot
```
```{r}
gatecheck()
```

Looking at the graph around 36, it looks like the AND & XOR are switched - so switching those back:

```{r}
gates<-gateswitcher("gfv","hcm",gates)


gatesdf<-as.data.frame(matrix(nrow=0,ncol=4))
for(i in 1:length(gates)){gatesdf<-rbind(gatesdf,gates[[i]])}
colnames(gatesdf)<-c("a","op","b","c")
```



For some reason that I can't understand, z is wrong unless I run this several times.  That does not make sense.  And yet, it's happened multiple times.
```{r,echo=FALSE,warning=FALSE}
### recalculate z

actualzs<-getzs(wirevalues,gates,highz)
zvals<-actualzs[[1]]
zdec<-actualzs[[2]]
actualzs<-getzs(wirevalues,gates,highz)
zvals<-actualzs[[1]]
zdec<-actualzs[[2]]
actualzs<-getzs(wirevalues,gates,highz)
zvals<-actualzs[[1]]
zdec<-actualzs[[2]]


### re-compare the add & redo the graph

cat("x=",xdec,"\n")
cat("y=",ydec,"\n")
cat("x+y=",x_plus_y,"\n")
cat("z=",zdec,"\n")

knitr::kable(rbind(0:45,xs=c(xvals,0),ys=c(yvals,0),added=c(x_plus_y_bin),zs=zvals))



#### graph of the gates
gatemat<-matrix(ncol=2,nrow=0)
for(i in 1:length(gates)){
  g<-gates[[i]]
  gatemat<-rbind(gatemat,c(g[1],g[4]),c(g[3],g[4]))}
gategraph<-graph_from_edgelist(gatemat)
ea<-c()
for(i in 1:length(gates)){ea<-c(ea,gates[[i]][2],gates[[i]][2])}
edge_attr(gategraph)<-list(op=ea)
gategraphplot<-ggraph(gategraph,layout="mds")+
  geom_edge_fan(aes(color=op),
                arrow=arrow(length= unit(1.5, 'mm')))+
  geom_node_label(aes(label=name),size=1.5,repel=TRUE)
gategraphplot
```




