---
title: "Day 7 Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
library(reshape2)
library(knitr)
library(dplyr)
library(stringr)
library(tidyverse)
library(readr)
library(collections)
options(scipen = 999)
```

```{r}
input<-read_lines("Day7Sample.txt")

cal_eqs<-lapply(str_split(input, " |: "),as.numeric)
### turn the equations into vectors
```

## Part 1
Checks to see if the test is valid - returns true if the equation can be made true 

```{r}
valid_eq<-function(eq){
  ### this probably doesn't need to be a priority queue - but what the heck.
  pq<-priority_queue()
  ### the target is the test value
  target<-eq[1]
  eq<-eq[-1]
  ### start with the assumption that it is invalid
  valid<-FALSE
  ### put the elements into the queue
  pq$push(eq,0)
  while(pq$size()>0){
    e<-pq$pop()
    ### if e has length 1 check to see if it equals the target
    if(length(e)==1){
      if(e==target){
        ### if so, there's a valid answer - stop
        valid<-TRUE
        break
        ### otherwise keep checking
      }else{next}}
    ### as all of the numbers are positive,
    ### if the first element is greater then target, no reason to continue
    if(e[1]>target){next}
    ### otherwise, add the first two elements and put the remainder into the queue
    eplus<-c(e[1]+e[2],e[-c(1,2)])
    ### or multiply them
    emult<-c(e[1]*e[2],e[-c(1,2)])
    pq$push(eplus,eplus[1])
    pq$push(emult,emult[1])}
  valid}

```

For part 1 - find the sum of all of the equations that can be made valid , 
```{r}

part1<-sum(sapply(cal_eqs,function(x){y<-valid_eq(x)*x[1]}))
part1
```


## Part 2

First pass - 
Just add another operator to the options

```{r}
valid_concat_eq<-function(eq){
  a<-Sys.time()
  ### this probably doesn't need to be a priority queue - but what the heck.
  pq<-priority_queue()
  ### the target is the test value
  target<-eq[1]
  eq<-eq[-1]
  ### start with the assumption that it is invalid
  valid<-FALSE
  ### put the elements into the queue
  pq$push(eq,0)
  while(pq$size()>0){
    e<-pq$pop()
    ### if e has length 1 check to see if it equals the target
    if(length(e)==1){
      if(e==target){
        ### if so, there's a valid answer - stop
        valid<-TRUE
        break
        ### otherwise keep checking
      }else{next}}
    ### as all of the numbers are positive,
    ### if the first element is greater then target, no reason to continue
    if(e[1]>target){next}
    ### otherwise, add the first two elements and put the remainder into the queue
    eplus<-c(e[1]+e[2],e[-c(1,2)])
    ### or multiply them
    emult<-c(e[1]*e[2],e[-c(1,2)])
    ### or concatenate them
    econcat<-c(as.numeric(str_c(e[1],e[2])),e[-c(1,2)])
    pq$push(eplus,eplus[1])
    pq$push(emult,emult[1])
    pq$push(econcat,econcat[1])}
  valid}

```

The same for part 2

```{r}
part2<-sum(sapply(cal_eqs,function(x){y<-valid_concat_eq(x)*x[1]}))
part2
```

This works - but it's slow.  The successes are pretty fast, but the failures (especially the large failures) take quite a while to finish failing.

The solutions thread suggested working backwards (r-to-l rather than l-to-r), which drops failing equations much, much, much more quickly.

```{r}
valid_concat_eq_back<-function(eq){
  ### this probably doesn't need to be a priority queue - but what the heck.
  pq<-priority_queue()
  ### start with the assumption that it is invalid
  valid<-FALSE
  ### put the elements into the queue
  pq$push(eq,0)
  while(pq$size()>0){
    e<-pq$pop()
    ### if e has length 2 check to see if it equals the target
    if(length(e)==2){
      if(e[1]==e[2]){
        ### if so, there's a valid answer - stop
        valid<-TRUE
        break
        ### otherwise keep checking
      }else{next}}
    ### as all of the numbers are positive and all operators can only make the number bigger
    ### if the first element is greater then target, no reason to continue
    if(e[2]>e[1]){next}
    ### used to find the last element of the equation
    leneq<-length(e)
    ### otherwise, subtract the last element from the target
    eplus<-c(e[1]-e[leneq],e[2:(leneq-1)])
    pq$push(eplus,-eplus[1])
    ### or check to see if it evenly divides into the target- if so, do it, if not, drop it
    if(e[1]%%e[leneq]==0){
      emult<-c(e[1]/e[leneq],e[2:(leneq-1)])
      pq$push(emult,-emult[1])}
    ### or check to see if the last element's digits match the target's final digits.
    ### if so, drop them; if not, drop it
    if(e[1]%%(10^nchar(e[leneq]))==e[leneq]){
    econcat<-c(e[1]%/%(10^nchar(e[leneq])),e[2:(leneq-1)])
    pq$push(econcat,-econcat[1])}
    }
  valid}
```

```{r}
part2<-sum(sapply(cal_eqs,function(x){y<-valid_concat_eq_back(x)*x[1]}))
part2
```
