---
title: "Day 15 Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
library(memoise)
library(ggplot2)
library(reshape2)
library(knitr)
library(dplyr)
library(stringr)
library(tidyverse)
library(readr)
library(collections)
options(scipen = 999)
```

```{r}
input<-read_lines("Day15Sample2.txt")
bl<-min(which(input==""))

### create the warehouse
warehouse<-matrix(nrow=(bl-1),ncol=nchar(input[1]))
for(i in 1:(bl-1)){warehouse[i,]<-unlist(str_split(input[i],""))}
### get the instructions
instructions<-c()
for(i in (bl+1):length(input)){
instructions<-c(instructions,unlist(str_split(input[i],"")))}
```

## Part 1

Creating a function - push()

This should take a vector with the wall at 1 & the robot at the end.  Will return list - first element is T/F if something could move.  Second is the updated, pushed vector.

```{r}
push<-function(v){
  wall<-max(which(v=="#"))
  if(!("."%in%v)){return(list(FALSE))}
  blnk<-max(which(v=="."))
  ### if there's no blanks before the wall, return FALSE
  if(blnk<wall){return(list(FALSE))}
  ### otherwise, return TRUE
list(TRUE,c(v[-blnk],"."))}
```


Move the robot
```{r}
rearrange<-function(w,inst){
  maxx<-ncol(w)
  maxy<-nrow(w)
  ### find the robot starting place
  r<-which(w=="@",arr.ind = TRUE)
  rx<-r[1,2]
  ry<-r[1,1]
  ### run through each instruction
  for(i in 1:length(inst)){
    nxti<-inst[i]
    switch(nxti,
           "^"={
#            cat("^",w[1:ry,rx],"\n")
             up<-push(w[1:ry,rx])
             ## if can't push, go to the next instruction
             if(up[[1]]==FALSE){next
               ### otherwise, move everything & update the robot's place
             }else{w[1:ry,rx]<-up[[2]]
             ry<-ry-1}
           },
           "<"={
 #            cat("<",w[ry,1:rx],"\n")
             left<-push(w[ry,1:rx])
             ## if can't push, go to the next instruction
             if(left[[1]]==FALSE){next
               ### otherwise, move everything & update the robot's place
             }else{w[ry,1:rx]<-left[[2]]
             rx<-rx-1}
           },
           "v"={
#             cat("v",w[maxy:ry,rx],"\n")
             down<-push(w[maxy:ry,rx])
             ## if can't push, go to the next instruction
             if(down[[1]]==FALSE){next
               ### otherwise, move everything & update the robot's place
             }else{w[maxy:ry,rx]<-down[[2]]
             ry<-ry+1}
           },
           ">"={
#             cat(">",w[ry,maxx:rx],"\n")
             right<-push(w[ry,maxx:rx])
             ## if can't push, go to the next instruction
             if(right[[1]]==FALSE){next
               ### otherwise, move everything & update the robot's place
             }else{w[ry,maxx:rx]<-right[[2]]
             rx<-rx+1}
           },
           cat("bad direction\n"))}
  ### calculate gps
  gps<-as.data.frame(which(w=="O",arr.ind=TRUE))
  gps<-gps%>%mutate(crd=100*(row-1)+(col-1))
sum(gps$crd)}
```



```{r}
part1<-rearrange(warehouse,instructions)
part1
```
## Part 2

First, double the width of the warehouse

```{r}
### create the warehouse
widehouse<-matrix(nrow=(bl-1),ncol=2*nchar(input[1]))
for(i in 1:(bl-1)){
  y<-unlist(str_split(input[i],""))
  y<-sapply(y,function(x){
    switch(x,
           "#"=c("#","#"),
           "O"=c("[","]"),
           "."=c(".","."),
           "@"=c("@","."),
           NA)})
  widehouse[i,]<-y}
```

Left and right do not change. Up and down, do
pushup() & pushdown return FALSE if a move is not possible. They return the robot and any box halves that need to move if it is possible.  For sorting purposes, they're in format y~x and padded with leading 0s to three digits.

```{r}
pushup<-function(w,x,y){
  chgs<-c()
  u<-w[y-1,x]
  switch(u,
         "."={
           chgs<-str_flatten(str_flatten(c(sprintf("%03d",y),sprintf("%03d",x)),"~"),"~")},
         "#"=return(FALSE),
         "["={
           above<-pushup(w,x,y-1)
           if(above[1]==FALSE){return(FALSE)}
           over<-pushup(w,x+1,y-1)
           if(over[1]==FALSE){return(FALSE)}
           chgs<-c(str_flatten(str_flatten(c(sprintf("%03d",y),sprintf("%03d",x)),"~"),"~"),above,over)},
         "]"={
           above<-pushup(w,x,y-1)
           if(above[1]==FALSE){return(FALSE)}
           over<-pushup(w,x-1,y-1)
           if(over[1]==FALSE){return(FALSE)}
           chgs<-c(str_flatten(str_flatten(c(sprintf("%03d",y),sprintf("%03d",x)),"~"),"~"),above,over)},
         cat("u=",u,"unknown\n"))
  unique(chgs)}

### I should have changed above to below - but I didn't.
pushdown<-function(w,x,y){
  chgs<-c()
  u<-w[y+1,x]
  switch(u,
         "."={
           chgs<-str_flatten(c(sprintf("%03d",y),sprintf("%03d",x)),"~")},
         "#"=return(FALSE),
         "["={
           above<-pushdown(w,x,y+1)
           if(above[1]==FALSE){return(FALSE)}
           over<-pushdown(w,x+1,y+1)
           if(over[1]==FALSE){return(FALSE)}
           chgs<-c(str_flatten(c(sprintf("%03d",y),sprintf("%03d",x)),"~"),above,over)},
         "]"={
           above<-pushdown(w,x,y+1)
           if(above[1]==FALSE){return(FALSE)}
           over<-pushdown(w,x-1,y+1)
           if(over[1]==FALSE){return(FALSE)}
           chgs<-c(str_flatten(c(sprintf("%03d",y),sprintf("%03d",x)),"~"),above,over)},
         cat("d=",u,"unknown\n"))
  unique(chgs)}
```


For up and down, the rearrangewide() sorts everything that moves and then moves it individually from top to bottom (for up) or bottom to top (for down) 

```{r}
rearrangewide<-function(w,inst){
  maxx<-ncol(w)
  maxy<-nrow(w)
  ### find the robot starting place
  r<-which(w=="@",arr.ind = TRUE)
  rx<-r[1,2]
  ry<-r[1,1]
  ### run through each instruction
  for(i in 1:length(inst)){
    nxti<-inst[i]
    switch(nxti,
           "^"={
#            cat("^",w[1:ry,rx],"\n")
             up<-pushup(w,rx,ry)
             ## if can't push, go to the next instruction
             if(up[1]==FALSE){next
               ### otherwise, move everything & update the robot's place
             }else{
               ### sort up from the top to the bottom
              up<-sort(up)
              for(j in 1:length(up)){
                ### get the coordinate
                u<-unlist(lapply(str_split(up[j],"~"),as.numeric))
                ### copy its value to the row above
                w[(u[1]-1),u[2]]<-w[u[1],u[2]]
                ### put a period there - will be overwritten if something is put into it
                w[u[1],u[2]]<-"."}
              w[ry,rx]<-"."
              ry<-ry-1}
           },
### left doesn't change
           "<"={
             left<-push(w[ry,1:rx])
             ## if can't push, go to the next instruction
             if(left[[1]]==FALSE){next
               ### otherwise, move everything & update the robot's place
             }else{w[ry,1:rx]<-left[[2]]
             rx<-rx-1}
           },
           "v"={
#             cat("v",w[maxy:ry,rx],"\n")
             down<-pushdown(w,rx,ry)
             ## if can't push, go to the next instruction
             if(down[1]==FALSE){next
               ### otherwise, move everything & update the robot's place
             }else{
              down<-sort(down,decreasing=TRUE)
              for(j in 1:length(down)){
                ### get the coordinate
                d<-unlist(lapply(str_split(down[j],"~"),as.numeric))
                ### copy its value to the row above
                w[(d[1]+1),d[2]]<-w[d[1],d[2]]
                w[d[1],d[2]]<-"."}
              w[ry,rx]<-"."
              ry<-ry+1}
           },
### right doesn't change
           ">"={
             right<-push(w[ry,maxx:rx])
             ## if can't push, go to the next instruction
             if(right[[1]]==FALSE){next
               ### otherwise, move everything & update the robot's place
             }else{w[ry,maxx:rx]<-right[[2]]
             rx<-rx+1}
           },
           cat("bad direction\n"))
    
    stillboxes<-which(w=="[")+nrow(w)
    if(any(w[stillboxes]!="]")){return(list(i,nxti,w,stillboxes))}
    stillboxes<-which(w=="]")-nrow(w)
    if(any(w[stillboxes]!="[")){return(list(i,nxti,w,stillboxes))}
    }
  ### calculate gps
  gps<-as.data.frame(which(w=="[",arr.ind=TRUE))
  gps<-gps%>%mutate(crd=100*(row-1)+(col-1))
sum(gps$crd)}
```

```{r}
part2<-rearrangewide(widehouse,instructions)
part2
```
