---
title: "Day 8 Notebook"
output: html_notebook
---

```{r setup, include=FALSE}
library(ggplot2)
library(reshape2)
library(knitr)
library(dplyr)
library(stringr)
library(tidyverse)
library(readr)
library(collections)
options(scipen = 999)
```

```{r}
input<-read_lines("Day8Sample.txt")
```
create a dictionary of the antennas.
also find the edges of the map

```{r}
antennadict<-dict()
for(i in 1:length(input)){
  x<-unlist(str_split(input[i],""))
  for(j in 1:length(x)){
    if(x[j]!="."){
      antennadict$set(x[j],c(antennadict$get(x[j],default=list()),list(c(j,i)) ) )}}}
maprows<-length(input)
mapcolumns<-nchar(input[1])
```


## Part 1

Quick function for finding anti-node pairs - takes in the antennas & map limits
```{r}
get_ans<-function(antenna1, antenna2, mr, mc){
  ans<-list()
  dif<-antenna1-antenna2
  an1<-antenna1+dif
  an2<-antenna2-dif
  ### if within bounds, it's a valid node
  if(all(an1>=c(1,1))&&all(an1<=c(mc,mr))){ans<-c(ans,list(an1))}
  if(all(an2>=c(1,1))&&all(an2<=c(mc,mr))){ans<-c(ans,list(an2))}
ans}

```

Find the map strength

```{r}
mapstrength<-function(ad,mr,mc){
  ### get the map frequency
  freqs<-unlist(ad$keys())
  ### create a list of antinodes
  all_antennas<-list()
  ### for each frequency, find all of the antinodes possible
  for(i in 1:length(freqs)){
    f<-freqs[i]
    f_antennas<-ad$get(f)
    if(length(f_antennas)>=2){
    for(j in 1:(length(f_antennas)-1)){
      for(k in (j+1):length(f_antennas)){
        all_antennas<-c(all_antennas,get_ans(f_antennas[[j]],f_antennas[[k]],mr,mc))}}}}
  ### get rid of duplicates
  all_antennas<-unique(all_antennas)
  all_antennas}
```

```{r}
### p1 list of all antinodes
p1<-mapstrength(antennadict,maprows,mapcolumns)
### part 1 is the length of the list
part1<-length(p1)
part1
```
## Part 2


Just change how to find anti-nodes so that it keeps going until it reaches the edge of the map and includes the antennas themselves


```{r}
get_harmonics<-function(antenna1, antenna2, mr, mc){
  ### add the antennas to the list
  ans<-c(list(antenna1),list(antenna2))
  dif<-antenna1-antenna2
  an1<-antenna1+dif
  ### keep adding dif to antenna1
  while(all(an1>=c(1,1))&&all(an1<=c(mc,mr))){
    ans<-c(ans,list(an1))
    an1<-an1+dif}
  ### do the same for an2
  an2<-antenna2-dif
  while(all(an2>=c(1,1))&&all(an2<=c(mc,mr))){
    ans<-c(ans,list(an2))
    an2<-an2-dif}
ans}

```

With the new map strength that includes harmonics:

```{r}
harmonicstrength<-function(ad,mr,mc){
  ### get the map frequency
  freqs<-unlist(ad$keys())
  ### create a list of antinodes
  all_antennas<-list()
  ### for each frequency, find all of the antinodes possible
  for(i in 1:length(freqs)){
    f<-freqs[i]
    f_antennas<-ad$get(f)
    ### only do this if there are at least 2 antennas of any given type
    if(length(f_antennas)>=2){
      for(j in 1:(length(f_antennas)-1)){
        for(k in (j+1):length(f_antennas)){
          all_antennas<-c(all_antennas,get_harmonics(f_antennas[[j]],f_antennas[[k]],mr,mc))}}}}
  ### get rid of duplicates
  all_antennas<-unique(all_antennas)
  all_antennas}
```

Rerun part 2.

```{r}
### p2 is the list of all antinodes
p2<-harmonicstrength(antennadict,maprows,mapcolumns)
### part 2 is the length of the list
part2<-length(p2)
part2
```


